---
title: "Mafalda-RBM25"
author: "Pedro"
date: "2024-09-25"
output: html_document
---

# Libraries
```{r}
options(repos = c(CRAN = "https://cloud.r-project.org/"))
install.packages(c("BiocManager", "remotes"))
BiocManager::install(version = "3.18", ask = FALSE)
bioCon <- c("QFeatures","msqrob2", "SingleCellExperiment", "scp", "scpdata", "scater", "Biostrings","org.Hs.eg.db", "GO.db", "destiny", "zellkonverter", "scran", "ComplexHeatmap", "bluster")
CRANpkgs <- c("tidyverse","viridis", "patchwork", "plotly", "impute", "RcolorBrewer", "limma", "circlize", "reticulate", "ggrepel", "ggpp", "gridExtra")

install.packages(CRANpkgs)
BiocManager::install(bioCon)
#BiocManager::install("UCLouvain-CBIO/scp#50")

## Core packages
library("scp")

## Utility packages
library("patchwork")
library("scater")
library("Biostrings")
library("destiny")
library("impute")
library("scran")
library("ComplexHeatmap")
library("RColorBrewer")
library("limma")
library("bluster")
library("circlize")
library("ggrepel")
library("ggpp")
library("gridExtra")
library("viridis")
library("plotly")
library("org.Hs.eg.db")
library("GO.db")
library("msqrob2")
library("reshape2")
library("tidyverse")
```

# Summary statistics of ms runs
```{r}
# General ID overview of the runs
dir <- "/Volumes/bio/MBT-g-Cell-Diversity-Lab/PEDRO/PorseLab/Mafalda/20240711_152750_Mafalda_cKIT"
  
# Fetch all run overview files and store summary statistics in stats
run_list <- list.files(dir, recursive = T, pattern = '_RunOverview.tsv')
stats <- data.frame(run_list)
stats <- stats %>% column_to_rownames(., var = 'run_list')

wt1_replicates <- c("1888", "1900", "1904", "1914")
wt2_replicates <- c("1866", "1880", "1894", "1912")
hetko_replicates <- c("1864", "1868", "1892", "1902")

# Combine all replicates into a single vector for matching
Replicate <- c(wt1_replicates, wt2_replicates, hetko_replicates)

# Create the regex pattern for replicates
replicate_pattern <- paste(Replicate, collapse = "|")

# Loop through each run file in the run list
for (r in run_list) {
  # Read the data from each run file
  prot <- read.delim(paste(dir, '/', r, sep = ""), skipNul = TRUE) %>% as_tibble()
  
  # Extract various statistics from the file
  stats[r, 'Precursors'] <- as.numeric(gsub(",", "", sub(" .*", "", prot$Value[1])))
  stats[r, 'Peptides'] <- as.numeric(gsub(",", "", sub(" .*", "", prot$Value[3])))
  stats[r, 'Protein.Groups'] <- as.numeric(gsub(",", "", sub(" .*", "", prot$Value[4])))
  stats[r, 'Data.Points.per.Peak.(MS1)'] <- str_extract(prot$Value[25], "[^ ]+")
  stats[r, 'Data.Points.per.peak.(MS2)'] <- str_extract(prot$Value[26], "[^ ]+")
  stats[r, 'Median.FWHM'] <- sub(",", ".", str_extract(prot$Value[21], "[^ ]+"))
  stats[r, 'Median.Peak.Width'] <- sub(",", ".", str_extract(prot$Value[22], "[^ ]+"))
  stats[r, '%.Missed.Cleavages'] <- sub(",", ".", str_extract(prot$Value[9], "[^ ]+"))
  stats[r, '%.Missed.Cleavages.Identified'] <- sub(",", ".", str_extract(prot$Value[10], "[^ ]+"))

  # Extract the replicate number from the filename using the pattern
  replicate <- str_extract(r, replicate_pattern)
  stats[r, "Replicate"] <- replicate

  # Assign the correct condition based on the replicate value
  if (!is.na(replicate)) {
    if (replicate %in% wt1_replicates) {
      stats[r, "Condition"] <- "WT 1: +/+; Mx1-Cre"
    } else if (replicate %in% wt2_replicates) {
      stats[r, "Condition"] <- "WT 2: fl/+"
    } else if (replicate %in% hetko_replicates) {
      stats[r, "Condition"] <- "HetKO: fl/+; Mx1-Cre"
    }
  } else {
    stats[r, "Condition"] <- NA  # Handle cases where the replicate does not match any known condition
  }
}

#Map the cell types from the filenames to the actual cell types
stats <- stats %>%
  rownames_to_column("filename") %>%
  mutate(filename = sub("\\.raw.*", "", filename)) %>% 
  mutate(filename = sub(".*RunSummaries/", "", filename))

#Plot the summary statistics of identifications
pivot.stats <- stats %>%
  pivot_longer(cols = c("Precursors", "Peptides", "Protein.Groups"), names_to = "Variable", values_to = "Value") %>%
  group_by(Replicate, Variable) %>%
  summarise(Mean = mean(Value, na.rm = TRUE)) %>%
  mutate(Variable = factor(Variable, levels = c("Precursors", "Peptides", "Protein.Groups")))

#Plot the means as bars and individual measurements as jitter
ggplot() +
geom_bar(data = pivot.stats, aes(x = Replicate, y = Mean, fill = reorder(Variable, -Mean)),
        position = "dodge", stat = "identity") +
geom_jitter(data = stats, aes(x = as.numeric(factor(Replicate)) - 0.275, y = Precursors, color = "Precursors"), 
            position = position_jitter(width = 0.07), size = 1, show.legend = FALSE) +
geom_jitter(data = stats, aes(x = as.numeric(factor(Replicate)), y = Peptides, color = "Peptides"), 
            position = position_jitter(width = 0.07), size = 1, show.legend = FALSE) +
geom_jitter(data = stats, aes(x = as.numeric(factor(Replicate)) + 0.275, y = Protein.Groups, color = "Protein.Groups"), 
            position = position_jitter(width = 0.07), size = 1, show.legend = FALSE) +
labs(x = "Mouse #", y = "Count", fill = "Feature") +
scale_fill_manual(values = c("#0077be","#92b0da", "#b7d1c8"), name = "Feature", labels = c("Precursors", "Peptides", "Protein Groups")) +
scale_color_manual(values = c("#635b83", "#a3c4f3","#6b9d9b")) +
scale_y_continuous(breaks = seq(0, 60000, by = 5000)) +
theme_bw() +
theme(axis.text.x = element_text(angle = 45, hjust = 1))
ggsave("20240711_152750_Mafalda_cKIT/Figures/summary_stats.pdf", width = 7.5, height = 5, dpi = 300)
```

# Intial clean up and data preparation
To accomodate the scp package requirements for analysis, the data was reshaped and the metadata was added to the quantitative data.
```{r}
SN.PEPquant <- read.delim("20240711_152750_Mafalda_cKIT/Mafalda_cKIT_Report_PEPQuant (Normal).tsv")

SN.PEPquant <- merge(SN.PEPquant, stats[, c("filename", "Condition", "Replicate")], 
                     by.x = "R.FileName", by.y = "filename", all.x = TRUE)

#reshape the data to generate the input table and sample table required for the scp package
inputCols <- c("FG.MS1Quantity", "R.FileName", "PG.Genes", "PG.ProteinAccessions", "PG.IsSingleHit", "PEP.StrippedSequence", "FG.Charge", "EG.Qvalue")
sampleCols <- c("R.FileName", "Condition", "Replicate")

sampleData <- SN.PEPquant %>% 
  select(all_of(sampleCols)) %>%
  distinct() %>%
  mutate(Quant.Col = "FG.MS1Quantity")

inputData <- SN.PEPquant %>%
  select(all_of(inputCols))
```

# Create a QFeatqures object
```{r}


names(sampleData)[1] <- "runCol" #This is a problem with .checkRunCol() function
scp <- readQFeaturesFromDIANN(assayData = inputData,
                            colData = sampleData,
                            runCol = "R.FileName",
                            quantCols ="FG.MS1Quantity",
                            removeEmptyCols = TRUE)


scp <- zeroIsNA(scp, i = names(scp))
scp <- infIsNA(scp,i = names(scp))


# Fixes typo on the colData
coldata <- colData(scp)
coldata$Condition[is.na(coldata$Condition)] <- "HetKO: fl/+; Mx1-Cre" # Replaces NA values in Condition column
coldata$Replicate[is.na(coldata$Replicate)] <- "1892" # Replaces NA values in Replicate column
colData(scp) <- coldata
```

## Aggregate Precursors into Peptides
```{r}
#Filter for Qval < 0.01
scp <- filterFeatures(scp,
                      ~ EG.Qvalue <= 0.01)

scp <- aggregateFeaturesOverAssays(scp,
                                  i = names(scp),    ## assays to be aggregated
                                  name = paste0("Peptides_", names(scp)), ## new assay name
                                  fcol = "PEP.StrippedSequence", 
                                  fun = matrixStats::colMedians,
                                  na.rm = TRUE)
scp <- joinAssays(scp,
                i = grep("Peptides_", names(scp)),
                name = "Peptides")

# Log2 transoform
scp <- logTransform(scp,
                    i = "Peptides",
                    name = "Peptides_log")

scp <- normalize(scp,
                    i= "Peptides_log",
                    name = "lognorm_Peptides",
                    method = "center.median")

# Normalize the data
# TODO: Keep an eye on other normalization methods
par(mfrow = c(1, 4))
limma::plotDensities(assay(scp[["Peptides"]]), legend = FALSE, main = "non-log")
limma::plotDensities(assay(scp[["Peptides_log"]]), legend = FALSE, main = "log2")
limma::plotDensities(assay(scp[["lognorm_Peptides"]]), legend = FALSE, main = "log2 normal")
boxplot(assay(scp[["lognorm_Peptides"]]),
    col = palette()[-1], ylab = "intensity",
    xlab = "samples",
    main = "log2 normal intensities",
    xaxt = "n")
plot(scp)
```
### Peptide Imputation
```{r}
# TODO: Figure out how to get rid of these bad precursors. 
# Impute with MinDet
scp <- impute(scp,
              i = "lognorm_Peptides",
              method = "MinDet",
              name = "Peptides_imp_mindet",
              MARGIN = 1L) #Performs imputation on a feature-wise basis

par(mfrow = c(1, 2))
limma::plotDensities(assay(scp[["lognorm_Peptides"]]), legend = FALSE, main = "none-imputed")
limma::plotDensities(assay(scp[["Peptides_imp_mindet"]]), legend = FALSE, main=  "minimal val - feature wise")
table(is.na(assay(scp[["lognorm_Peptides"]])))
table(is.na(assay(scp[["Peptides_imp_mindet"]])))
```

## Protein aggregation
```{r}
#This function calculates the robust summarisation for each feature (protein).
#Note that the function assumes that the intensities in input e are already log-transformed.
scp <- aggregateFeatures(scp,
                         i = "lognorm_Peptides",
                         fcol= "PG.ProteinAccessions",
                         name = "Proteins",
                         fun = MsCoreUtils::robustSummary) 

#aggregate imputed features as well
scp <- aggregateFeatures(scp,
                         i = "Peptides_imp_mindet" ,
                         fcol= "PG.ProteinAccessions",
                         name = "Proteins_imp",
                         fun = MsCoreUtils::robustSummary) 


#We now perform centering followed by scalling of the data to unit variance
means <- rowMeans(assay(scp, "Proteins"))
scp <- sweep(scp, MARGIN = 1, STATS = means, FUN = "-",
             i = "Proteins", name = "Proteins_centered")
stds <- rowSds(assay(scp, "Proteins_centered"))
scp <- sweep(scp, MARGIN = 1, STATS = stds, FUN = "/",
             i = "Proteins_centered", name = "Proteins_scaled")

par(mfrow = c(1, 3))
limma::plotDensities(assay(scp[["Proteins_scaled"]]), legend = FALSE, main = "Z-scored")
limma::plotDensities(assay(scp[["Proteins_centered"]]), legend = FALSE, main = "centered")
limma::plotDensities(assay(scp[["Proteins"]]), legend = FALSE, main = "'raw'")

plot(scp, interactive = TRUE) 
table(rowData(scp[["Proteins"]])$.n)
table(is.na(assay(scp[["Proteins"]])))

par(mfrow = c(1, 2))
limma::plotDensities(assay(scp[["Proteins"]]), legend = FALSE, main = "non-imputed")
limma::plotDensities(assay(scp[["Proteins_imp"]]), legend = FALSE, main = "imputed")

```

# RBM25 exploration
## Foldchange calculation
```{r, fig.width= 4, fig.height= 5}
rowData(scp[["Proteins"]])
table(rowData(scp[["Proteins"]])$.n)

# Define the features of interest
features <- c("B2RY56")
feature_names <- c("B2RY56" = "rbm25")

# Loop through each feature and generate plots
# This was initially made to extract a vector of features of interest
for (feature in features) {
  
  # Subset and transform the data for the current feature
  feature_df <- longFormat(scp[feature, ]) |>
      as.data.frame() |>
      filter(assay %in% c("Proteins")) |>
      mutate(
      assay = factor(assay, levels = c("Proteins")),
      replicate = str_extract(primary, replicate_pattern),
      replicate = if_else(is.na(replicate), "1892", replicate),
      sample = str_extract(colname, "(?<=_)[0-9]+(?=FG\\.MS1Quantity)"),
      id = paste(replicate, sample, sep = "_"),
      condition = case_when(
        replicate %in% wt1_replicates ~ "WT 1: +/+; Mx1-Cre",
        replicate %in% wt2_replicates ~ "WT 2: fl/+",
        replicate %in% hetko_replicates ~ "HetKO: fl/+; Mx1-Cre"),
      feature_name = feature_names[feature]) |>
      filter(replicate != "1868")
}

# Calculate mean values for each condition
mean_values <- feature_df %>%
  group_by(condition) %>%
  summarise(mean_value = mean(2^value))

# Extract the mean value for HetKO condition (to be used for fold change calculations)
hetko_mean <- mean_values %>% filter(condition == "HetKO: fl/+; Mx1-Cre") %>% pull(mean_value)

# Calculate the fold change relative to HetKO
mean_values <- mean_values %>%
  mutate(fold_change = mean_value / hetko_mean)

# Create the barplot with annotations for fold change
ggplot(feature_df, aes(x = condition, y = 2^value)) +
  geom_bar(stat = "summary", fun = "mean", aes(fill = condition)) +
  geom_jitter(aes(color = condition), width = 0.15, size = 2, alpha = 0.7) +
  stat_summary(fun.data = "mean_se", geom = "errorbar", width = 0.2, color = "black") +
  geom_text(data = mean_values %>% filter(condition != "HetKO: fl/+; Mx1-Cre"), 
            aes(x = condition, y = mean_value, label = paste0("FC: ", round(fold_change, 2))), 
            vjust = -1.5, size = 5, color = "black") +
  labs(x = "Condition", y = "Norm. expression") +
  theme_bw() +
  scale_fill_manual(values = c("WT 1: +/+; Mx1-Cre" = "#b3cde0",
                               "WT 2: fl/+" = "#006666",
                               "HetKO: fl/+; Mx1-Cre" = "#3d85c6")) +
  scale_color_manual(values = c("WT 1: +/+; Mx1-Cre" = "#8fa4b3",
                               "WT 2: fl/+" = "#7fb2b2",
                               "HetKO: fl/+; Mx1-Cre" = "#1e4263")) +
  theme(legend.position = "none",
        axis.text.x = element_text(angle = 45, hjust = 1))
ggsave("20240711_152750_Mafalda_cKIT/Figures/fold_change.pdf", width = 2, height = 4, dpi = 300)
```

## Expression profiles of RBM25
```{r, fig.width= 10, fig.height=7}
# Define a color palette for 12 replicates
replicate_colors <-  c("#6aa84f", "#d98e47", "#3d85c6", "#bf63bf", "#e89691",
                                "#FF1493", "#20B2AA", "#FF4500", "#b3cde0", "#006666", "#a7adba", "#FF69B4")

# Loop through each feature and generate plots
output_directory <- "20240711_152750_Mafalda_cKIT/Figures"

# Create the directory if it doesn't exist
if (!dir.exists(output_directory)) {
  dir.create(output_directory, recursive = TRUE)
}

for (feature in features) {
  
  # Subset and transform the data for the current feature
  feature_df <- longFormat(scp[feature, ]) |>
      as.data.frame() |>
      filter(assay %in% c("lognorm_Precursors","Peptides", "Proteins_scaled")) |>
      mutate(
      assay = factor(assay, levels = c("lognorm_Precursors", "Peptides", "Proteins_scaled")),
      replicate = str_extract(primary, replicate_pattern),
      replicate = if_else(is.na(replicate), "1892", replicate),
      sample = str_extract(colname, "(?<=_)[0-9]+$"),
      id = paste(replicate, sample, sep = "_"),
      condition = case_when(
        replicate %in% wt1_replicates ~ "WT 1: +/+; Mx1-Cre",
        replicate %in% wt2_replicates ~ "WT 2: fl/+",
        replicate %in% hetko_replicates ~ "HetKO: fl/+; Mx1-Cre"),
      feature_name = feature_names[feature]) 
  
  feature_df_mean <- feature_df %>%
    group_by(assay, replicate, sample, condition) %>%
    summarize(mean_value = mean(value, na.rm = TRUE), .groups = 'drop')

  # Plotting the original data points and the summarized mean lines on top
  p <- ggplot() +
       geom_point(data = feature_df, aes(x = sample, 
                                         y = value, 
                                         color = replicate, 
                                         group = replicate)) +
       geom_line(data = feature_df_mean, aes(x = sample, 
                                             y = mean_value, 
                                             color = replicate, 
                                             group = replicate),
                                             linetype = "solid", 
                                             size = 1) +  
       geom_point(data = feature_df_mean, aes(x = sample, 
                                              y = mean_value, 
                                              color = replicate, 
                                              group = replicate),size = 1) +
        facet_grid(assay ~ condition, scales = "free_y", drop = TRUE) +  
        scale_color_manual(values = replicate_colors) +  
        theme(axis.text.x = element_text(angle = 45, hjust = 1, size = 8)) +
        labs(title = paste("Feature:", feature_names[feature]))

  # Print the plot
  print(p)

  # Save the plot (uncomment to use)
  ggsave(
     filename = paste0(output_directory, "/", feature_names[feature], ".pdf"), 
     plot = p,  
     device = "pdf", 
     width = 8,       
     height = 6      
   )
}

# Fast access to .n
row_data <- as.data.frame(rowData(scp[["Proteins_scaled"]]))


# Heatmap
# Initialize an empty dataframe to store results
scaled_proteins <- data.frame()

# Loop through each feature to subset the data and filter for Proteins_scaled assay
for (feature in features) {
  
  # Subset and transform the data for the current feature, filtering for "Proteins_scaled"
  feature_df <- longFormat(scp[feature, ]) |>
      as.data.frame() |>
      filter(assay %in% c("Proteins_scaled")) |>
      mutate(
      assay = factor(assay, levels = c("Proteins_scaled")),
      replicate = str_extract(primary, replicate_pattern),
      replicate = if_else(is.na(replicate), "1892", replicate),
      sample = str_extract(colname, "(?<=_)[0-9]+$"),
      id = paste(replicate, sample, sep = "_"),
      condition = case_when(
        replicate %in% wt1_replicates ~ "WT 1: +/+; Mx1-Cre",
        replicate %in% wt2_replicates ~ "WT 2: fl/+",
        replicate %in% hetko_replicates ~ "HetKO: fl/+; Mx1-Cre"),
           feature_name = feature_names[feature])  # Add feature name for tracking
  
  # Append to the combined dataframe
  scaled_proteins <- rbind(scaled_proteins, feature_df)
}

scaled_proteins <- scaled_proteins %>%
  arrange(condition, replicate, sample) %>%
  mutate(
    position = as.numeric(factor(id, levels = unique(id))),  # Create a unique numeric position for each sample
    # Add spacing between conditions by incrementing positions for the next condition
    position = position + cumsum(!duplicated(condition)) - 1  # Adds a gap at the start of each new condition
  )

# Calculate the midpoint for each condition block to place the condition label
condition_labels <- scaled_proteins %>%
  group_by(condition) %>%
  summarize(
    midpoint = mean(range(position)),  # Calculate the midpoint of positions for the condition
    max_y = max(as.numeric(factor(feature_name)))  # Determine the top y position for placing the label
  )


# Generate the heatmap with square tiles
heatmap_expression <- ggplot(scaled_proteins, 
                             aes(x = position, y = feature_name, fill = value)) +
  geom_tile(width = 0.9, height = 0.9) +  # This creates the tiles for the heatmap
  scale_fill_distiller(palette = "RdBu", direction = -1,  # Reverse RdBu color palette
                       name = "Value", 
                       limits = c(-2.5, 2.5),  # Cap the gradient scale at -2.5 and 2.5
                       oob = scales::squish) +
  coord_fixed(ratio = 1) +  # Force a fixed aspect ratio of 1:1 for square tiles
  theme_minimal() +  # Minimal theme for a cleaner look
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),  # Rotate the x-axis labels
    panel.grid = element_blank()  # Remove the background grid lines
  ) +
  labs(x = "Samples", y = "Proteins") +
  scale_x_continuous(breaks = scaled_proteins$position, 
                     labels = scaled_proteins$id,  # Show the original sample IDs as labels
                     expand = expansion(mult = c(0.01, 0.05))) +  # Slight expansion to improve visual spacing
  expand_limits(y = max(condition_labels$max_y) + 2) +  # Expand y-axis to make room for condition labels
  # Add condition labels on top of the blocks
  geom_text(data = condition_labels, 
            aes(x = midpoint, y = max_y + 1, label = condition),  # Adjust position to move text higher
            size = 3, fontface = "bold", vjust = 0, 
            inherit.aes = FALSE)  # Prevent inheritance of unwanted aesthetics

# Print the heatmap
print(heatmap_expression)
ggsave("20240711_152750_Mafalda_cKIT/Figures/heatmap_expression.pdf", heatmap_expression, width = 10, height = 2, dpi = 300)
```

### List of Peptides in RBM25
```{r}
# Initialize an empty data frame to collect results across features
all_peptides_by_condition <- data.frame()

# Loop through each feature
for (feature in features) {
  
  # Subset and transform the data for the current feature
  feature_df <- longFormat(scp[feature, ]) |>
    as.data.frame() |>
    filter(assay %in% c("Peptides")) |>
    mutate(
      replicate = str_extract(primary, replicate_pattern),
      replicate = if_else(is.na(replicate), "1892", replicate),
      sample = str_extract(colname, "(?<=_)[0-9]+(?=FG\\.MS1Quantity)"),
      id = paste(replicate, sample, sep = "_"),
      condition = case_when(
        replicate %in% wt1_replicates ~ "WT 1: +/+; Mx1-Cre",
        replicate %in% wt2_replicates ~ "WT 2: fl/+",
        replicate %in% hetko_replicates ~ "HetKO: fl/+; Mx1-Cre"
      ),
      feature_name = feature_names[feature]
    )

  # Group by condition and collect unique peptide sequences from the 'rowname' column
  peptides_by_condition <- feature_df %>%
    group_by(condition) %>%
    summarize(peptides = list(unique(rowname)), .groups = 'drop')  # List unique peptides for each condition

  # Convert the list of peptides into a readable format if needed (optional)
  peptides_table <- peptides_by_condition %>%
    mutate(peptides = sapply(peptides, function(x) paste(x, collapse = ", ")))  # Concatenate peptide sequences

  # Append the results to the combined dataframe
  all_peptides_by_condition <- bind_rows(all_peptides_by_condition, peptides_table)
}

# Display the combined table of peptides by condition
view(all_peptides_by_condition)
```

# Missing values and predicted coverage
```{r, fig.width= 7.5, fig.height= 10}}
isIdentified <- function(x) colSums(!is.na(x)) != 0

scpNA <- aggregateFeaturesOverAssays(scp,
                                  i = 1:48,    ## assays to be aggregated
                                  name = paste0("MV_proteins", grep("20240705", names(scp)[1:48])), ## new assay name
                                  fcol = "PG.ProteinAccessions",    ## what to aggregate
                                  fun = isIdentified)                                  

scpNA <- joinAssays(scpNA,
                    i = grep("MV_proteins", names(scpNA)), 
                        name = "prot_id") 

isMissing_prot <- is.na(assay(scpNA[["prot_id"]]))
assay(scpNA[["prot_id"]])[isMissing_prot] <- FALSE

coldata <- colData(scpNA)
coldata$Condition[is.na(coldata$Condition)] <- "HetKO: fl/+; Mx1-Cre" # Replaces NA values in Condition column
coldata$Replicate[is.na(coldata$Replicate)] <- "1892" # Replaces NA values in Replicate column
colData(scpNA) <- coldata
#Here sensitivity is defined as the number of distinct peptide identifications.
#Local sensitivity reported as the number of features per cell.
#Total sensitivity reported as the number of features across all cells within a celltype.
prot_sensitivity_summary <- reportMissingValues(scpNA, "prot_id", by = scpNA$Replicate)
prot_sensitivity_summary <- rownames_to_column(prot_sensitivity_summary, var = "Replicate")

p1 <- ggplot(prot_sensitivity_summary, aes(x = Completeness*100, y = LocalSensitivityMean, color = Replicate, size = NumberCells)) +
        geom_point() +
        geom_linerange(aes(ymin = LocalSensitivityMean - LocalSensitivitySd, ymax = LocalSensitivityMean + LocalSensitivitySd), size = 1) +
        scale_color_manual("Mice",
                        values = c("#6aa84f", "#d98e47", "#3d85c6", "#bf63bf", "#e89691",
                                "#FF1493", "#20B2AA", "#FF4500", "#b3cde0", "#006666", "#a7adba", "#FF69B4"),
                        name = "Mouse #") +
        labs(x = "Completeness [%]", y = "Average protein features per Mouse", size = "Number of replicates", color = "Population") +
        theme_bw()
  

#Evaluate similarity/consistency of identifications across celltypes
ji_prot <- jaccardIndex(scpNA, "prot_id", by = scpNA$Condition)

medians <- ji_prot %>%
  group_by(by) %>%
  summarise(median = median(jaccard))

p2 <- ggplot(ji_prot, aes(x = jaccard, fill = by)) +
  geom_histogram(alpha = 0.7) +
  geom_vline(data = medians, aes(xintercept = median), linetype = "dashed", color = "#5b5b5b") +
  geom_text_repel(data = medians, aes(x = median, y = 0, label = paste("Median:", round(median, 2))), nudge_y = 0.5) +  
  facet_grid(by ~ ., scales = "free_y") +
  scale_fill_manual("Condition",
                    values = c("#3d85c6",  "#b3cde0", "#006666")) +
  labs(x = "Jaccard Index", y = "Count") +
  theme_minimal() +
  theme(strip.text = element_blank())

#TODO: Add correct colData information
csc_prot <- cumulativeSensitivityCurve(scpNA, "prot_id", by = scpNA$Condition,
                                  niters = 10, 
                                  nsteps = max(48))

pCSC_prot <- ggplot(csc_prot) +
  aes(x = SampleSize, y = Sensitivity, colour = by) +
  geom_point(size = 2, alpha = 0.7) +
  scale_color_manual("Condition",
                  values = c("#3d85c6",  "#b3cde0", "#006666")) + 
                   labs(x = "Number of samples", y = "Sensitivity") +
                   theme_minimal() 

predCSC_prot <- predictSensitivity(csc_prot, nSample = 1:length(unique(csc_prot$SampleSize)))

#Add the maximal predicted sensitivity to the plot
prot_maxSensitivity <- predictSensitivity(csc_prot, nSamples = Inf)

prot_maxSensitivity <- prot_maxSensitivity %>%
  dplyr::select(-SampleSize) %>%
  rename(Condition = by, "Max. predicted sensitivity" = Sensitivity) %>%
  dplyr::select(Condition, "Max. predicted sensitivity", everything()) %>%
  mutate(Condition = case_when(
    Condition == "HetKO: fl/+; Mx1-Cre" ~ "HetKO: fl/+; Mx1-Cre",
    Condition == "WT 2: fl/+" ~ "WT 2: fl/+",
    Condition == "WT 1: +/+; Mx1-Cre"  ~ "WT 1: +/+; Mx1-Cre"))

p3 <- pCSC_prot + 
        geom_line(data = predCSC_prot, size = 1) +
        geom_table(data = prot_maxSensitivity, 
                  aes(x = Inf, y = -Inf, label = list(prot_maxSensitivity)), 
                  hjust = 1, vjust = 0) +
                  theme(legend.position = "none") 
(p1 | p2) / p3 + plot_annotation(tag_levels = 'A') + plot_layout(guides = 'collect')
#ggsave("20240711_152750_Mafalda_cKIT/Figures/missing_values.pdf", width = 7.5, height = 7)
```

# Differential expression analysis using robust estimation 
## Clean up certain atrributes in ColData
```{r}
# Removes unsuccesfull KO
colData(scp) <- colData(scp)[colData(scp)$Replicate != "1868", ]
scp <- scp[, colnames(scp) %in% rownames(colData(scp))]

# Recode the condition column to deal wird rare expression patterns
colData(scp)$ab.condition <- dplyr::recode(colData(scp)$Condition,
                                           "HetKO: fl/+; Mx1-Cre" = "KO",
                                           "WT 1: +/+; Mx1-Cre" = "WT1",
                                           "WT 2: fl/+" = "WT2")
pe <- scp
```

## Robust parameter estimation
```{r}
# Estimation using msqrob
pe <- msqrob(object = pe, i = "Proteins", formula = ~ab.condition, overwrite = TRUE)

# Extract the coefficients
getCoef(rowData(pe[["Proteins"]])$msqrobModels[[1]])
# Make contrasts 
L <- makeContrast(c("ab.conditionWT1=0", 
                    "ab.conditionWT2=0"),
                  parameterNames = c(
                      "(Intercept)",
                      "ab.conditionWT1",
                      "ab.conditionWT2"))

#test hypothesis
pe <- hypothesisTest(object = pe, i = "Proteins", contrast = L, overwrite = TRUE)
```

### plot volcano plots
```{r, fig.width= 10, fig.height= 5}
volcanoMed <- rowData(pe[["Proteins"]])$ab.conditionWT1 %>%
    ggplot(aes(x = logFC, y = -log10(pval), color = adjPval < 0.01)) +
    geom_point(cex = 2.5) +
    scale_color_manual(values = alpha(c("black", "red"), 0.5)) +
    theme_minimal() +
    geom_vline(xintercept = log2(0.74 / .25), col = "red") +
    ggtitle("median summarisation")


WT1_KO <- rowData(pe[["Proteins"]])$ab.conditionWT1 %>% 
  rownames_to_column("Proteins") %>% 
  mutate(
    color = case_when(
      logFC > 0.5 & adjPval < 0.05 ~ "#9fbadf",   # Higher logFC and significant
      logFC < -0.5 & adjPval < 0.05 ~ "#ff661a",   # Lower logFC and significant
      TRUE ~ "#dcdfe1"                            # Everything else
    )
  )

top_proteins <- WT1_KO %>%
  filter(adjPval < 0.05 & abs(logFC) > 0.5) %>%
  top_n(25, wt = abs(logFC)) %>%
  bind_rows(WT1_KO %>% filter(Proteins == "B2RY56"))




#Heatmap on differentially expressed proteins
zscore_normalized <- t(apply(assay(pe[["Proteins"]]), 1, function(x) {
  (x - mean(x, na.rm = TRUE)) / sd(x, na.rm = TRUE)
}))

# Step 2: Subset the Z-score normalized data for the selected proteins (top_proteins)
subset_zscore <- zscore_normalized[rownames(zscore_normalized) %in% top_proteins$Proteins, ]

# Step 3: Convert the subsetted Z-score normalized matrix into long format
melted_data <- melt(subset_zscore)

# Step 4: Add sample information from colData to the melted data
coldata <- colData(pe) %>%
  as.data.frame() %>%
  rownames_to_column("Samples")

# Join with the melted data and clean up column names
melted_data <- melted_data %>%
  rename(PG.ProteinAccessions = Var1) %>%  # Rename Var1 to match the reference column
  left_join(data.frame(
    PG.ProteinAccessions = rowData(scp[["Proteins"]])$PG.ProteinAccessions,
    PG.Genes = rowData(scp[["Proteins"]])$PG.Genes
  ), by = "PG.ProteinAccessions") %>%  # Map accessions to gene symbols
  left_join(coldata, by = c("Var2" = "Samples")) %>%
  select(-Var2) %>%
  mutate(id = str_remove(runCol, ".*20ng_"))

# Create the heatmap using geom_tile, displaying NAs as a specific color (e.g., grey)
ggplot(melted_data, aes(x = id, y = PG.Genes, fill = value)) +
  geom_tile(width = 0.9, height = 0.9) +  # Create tiles for the heatmap
  scale_fill_distiller(palette = "RdBu", direction = -1,  # Reverse RdBu color palette
                       name = "Z-score", 
                       limits = c(-3, 3),  # Cap the gradient scale at -3 and 3
                       oob = scales::squish) +  # Squish out-of-bound values to the color limits
  facet_wrap(~ Condition, scales = "free_x", nrow = 1) +  # Segment by Condition
  theme_minimal() +
  theme(
    axis.text.x = element_text(angle = 45, hjust = 1, size = 8),  # Rotate the x-axis labels
    panel.grid = element_blank(),
    strip.text = element_text(size = 12, face = "bold")  # Customize condition block labels
  ) +
  labs(x = "Samples", y = "Proteins")
ggsave("20240711_152750_Mafalda_cKIT/Figures/DE_expression_heatmap.pdf", width = 10, height = 7)
```

#Principal Component Analysis
```{r, fig.width= 10, fig.height= 5}}
p <- pca(assay(scp, "Proteins_imp"), metadata = colData(scp), removeVar = 0.1)

screeplot(p, axisLabSize = 10, titleLabSize = 22)
#By condition
cond<-biplot(
  p,
  x = "PC1",
  y = "PC2",
  showLoadings = FALSE,
  ntopLoadings = 5,
  colby = "Condition",  # Color points by Condition
  colLegendTitle = "Condition",  # Legend title for Condition
  legendPosition = "right",  # Position the legend on the right
  pointSize = 3,
  legendLabSize = 12,
  legendTitleSize = 14,
  lab = NULL)

#By Mouse
mouse<-biplot(
  p,
  x = "PC1",
  y = "PC2",
  showLoadings = FALSE,
  ntopLoadings = 5,
  colby = "Replicate",  # Color points by Replicate
  colLegendTitle = "Mouse #",  # Legend title for Replicate
  legendPosition = "right",  # Position the legend on the right
  pointSize = 3,
  legendLabSize = 12,
  legendTitleSize = 14,
  lab = NULL,
  colkey = replicate_colors)

mouse + cond +  plot_layout(guides = 'collect')

```

